Unique solutions of contractions, CCS, and their HOL formalisation\thanksref{ATL1}

### 2020-07-31 ic-journal/equa.tex
Editorial changes to LaTeX sources

\begin{definition}[solutions and the unique solution]
  \label{def:solution}
Suppose $\{ X_i = E_i\}_{i\in I}$ is a system of equations: 
\begin{itemize}
\item
 $\til P$ is a \emph{solution of the system of equations (for $\wb$)} 
if for each $i$ it holds that $P_i \wb E_i [\til P]$;
\item The system has \emph{a unique solution for $\wb$}  if whenever 
 $\til P$ and $\til Q$ are both solutions then $\til P \wb \til Q$. 
\end{itemize} 
 \end{definition}
Similarily, the \emph{(unique) solution of a system of equations for $\sim$}
(or for $\rapprox$) can be obtained by replacing all occurrences of $\wb$
in the above definition with $\sim$ and $\rapprox$, respectively.

>>>>

\begin{definition}[solutions and the unique solution]
  \label{def:solution}
Suppose $\{ X_i = E_i\}_{i\in I}$ is a system of equations: 
\begin{itemize}
\item
 $\til P$ is a \emph{solution of the system of equations (for $\wb$)} 
if for each $i$ it holds that $P_i \wb E_i [\til P]$;
\item The system has \emph{a unique solution for $\wb$}  if whenever 
 $\til P$ and $\til Q$ are both solutions then $\til P \wb \til Q$. 
\end{itemize} 
 \end{definition}
 Similarly, % I&C editor issue 4
the \emph{(unique) solution of a system of equations for $\sim$}
(or for $\rapprox$) can be obtained by replacing all occurrences of $\wb$
in the above definition with $\sim$ and $\rapprox$, respectively.

### 2020-07-31 ic-journal/expansion.tex
Editorial changes to LaTeX sources

One basic property of rooted bisimiarity is congruence,
indeed it is  the coarsest congruence contained in bisimilarity
(see Section~\ref{s:coarsest} and Section~\ref{ss:context} for more details):
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

>>>>

One basic property of rooted bisimilarity is congruence, % editorial
indeed it is  the coarsest congruence contained in bisimilarity
(see Section~\ref{s:coarsest} and Section~\ref{ss:context} for more details):
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

### 2020-07-31 ic-journal/intro.tex
Editorial changes to LaTeX sources

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
Such structural simililarities can be found, for instance, in the
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that use the contraction preorder as an
auxiliary relation, and other unique-solution theorems (e.g. the one
for rooted bisimilarity 
 in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the ``up-to'' techniques).
%
In these cases, when moving between proofs there are only a few places in
the HOL proof scripts that have to be modified.
Then the successful termination of a proof gives us the
 guarantee that the proof is correct, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

@@

In Section~\ref{ss:ccs} we recall the core theory of CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Then, Section~\ref{s:eq} discusses equations and contractions, and in particular,
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
In Section~\ref{s:for} we highlight the CCS formalisation in HOL4, with
the unique-solution theorems in the \univariate case.
Section~\ref{sec:multivariate} describes the extension to the \multivariate case.
Finally, in Section~\ref{s:rel} and \ref{s:concl} we discuss the related work,
conclusions, and a few directions for future work.

>>>>

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
Such structural similarities can be found, for instance, in the % editorial changes
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that use the contraction preorder as an
auxiliary relation, and other unique-solution theorems (e.g. the one
for rooted bisimilarity 
 in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the ``up-to'' techniques).
%
In these cases, when moving between proofs there are only a few places in
the HOL proof scripts that have to be modified.
Then the successful termination of a proof gives us the
 guarantee that the proof is correct, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

@@

In Section~\ref{ss:ccs} we recall the core theory of CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Then, Section~\ref{s:eq} discusses equations and contractions, and in particular,
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
In Section~\ref{s:for} we highlight the CCS formalisation in HOL4, with
the unique-solution theorems in the \univariate case.
Section~\ref{sec:multivariate} describes the extension to the \multivariate case.
Finally, in Sections~\ref{s:rel} and \ref{s:concl} we discuss the related work, % editorial
conclusions, and a few directions for future work.

### 2020-07-31 ic-journal/related.tex
Editorial changes to LaTeX sources

Bengtson, Parrow and Weber did a substantial formalisation work
on CCS, $\pi$-calculi
and $\psi$-calculi 
using Isabelle/HOL and its nominal logic, with the main focus on the handling of
name binders \cite{bengtson2007completeness,parrow2009formalising}.
More details can be found in Bengtson's PhD thesis~\cite{bengtson2010formalising}. For CCS, 
he has formalized basic properties for strong/weak equivalence (congruence, basic
 algebraic laws); the CCS syntax does not have constants
or recursion, using instead replication.
% The formalisation effort has then been continued 
%
% On the other side, Jesper
% Bengtson and Joachim Parrow made great progress on $\pi$-calculus
% formalization and 
% proved that the algebraic axiomatization of bisimulation
% equivalence in the $\pi$-calculi is sound and
% complete. \cite{bengtson2007completeness} 
Other formalisations in this area include the early work of T.F.~Melham
\cite{melham1994mechanized} and O.A.~Mohamed
\cite{mohamed1995mechanizing} in HOL, Compton
\cite{compton2005embedding} in Isabelle/HOL,
Solange\footnote{\url{https://github.com/coq-contribs/ccs}} in Coq
and Chaudhuri et al.\;\cite{chaudhuri2015lightweight} in Abella, the latter
focuses on ``bisimulation up-to'' techniques (for strong bisimilarity)
for CCS and $\pi$-calculus.
Damien Pous \cite{pous2007new} also formalised up-to techniques and some CCS examples in
Coq.
Formalisations less related to ours
include Kahsai and Miculan \cite{kahsai2008implementing} for the spi
calculus in Isabelle/HOL, and Hirschkoff \cite{hirschkoff1997full} for the $\pi$-calculus in Coq.

>>>>

Bengtson, Parrow and Weber did a substantial formalisation work
on CCS, $\pi$-calculi
and $\psi$-calculi 
using Isabelle/HOL and its nominal logic, with the main focus on the handling of
name binders \cite{bengtson2007completeness,parrow2009formalising}.
More details can be found in Bengtson's PhD thesis~\cite{bengtson2010formalising}. For CCS, 
he has formalised basic properties for strong/weak equivalence (congruence, basic % editorial
 algebraic laws); the CCS syntax does not have constants
or recursion, using instead replication.
% The formalisation effort has then been continued 
%
% On the other side, Jesper
% Bengtson and Joachim Parrow made great progress on $\pi$-calculus
% formalization and 
% proved that the algebraic axiomatization of bisimulation
% equivalence in the $\pi$-calculi is sound and
% complete. \cite{bengtson2007completeness} 
Other formalisations in this area include the early work of T.F.~Melham
\cite{melham1994mechanized} and O.A.~Mohamed
\cite{mohamed1995mechanizing} in HOL, Compton
\cite{compton2005embedding} in Isabelle/HOL,
Solange\footnote{\url{https://github.com/coq-contribs/ccs}} in Coq
and Chaudhuri et al.\;\cite{chaudhuri2015lightweight} in Abella, the latter
focuses on ``bisimulation up-to'' techniques (for strong bisimilarity)
for CCS and $\pi$-calculus.
Damien Pous \cite{pous2007new} also formalised up-to techniques and some CCS examples in
Coq.
Formalisations less related to ours
include Kahsai and Miculan \cite{kahsai2008implementing} for the spi
calculus in Isabelle/HOL, and Hirschkoff \cite{hirschkoff1997full} for the $\pi$-calculus in Coq.

### 2020-07-31 ic-journal/unique.tex
Editorial changes to LaTeX sources

For this new relation, the analogous of Lemma~\ref{l:uptocon} and of
Theorem~\ref{t:contraBisimulationU} can now be stated without
constraints on summation.
The schema of the proofs is almost identical, because all 
properties of $\rcontr$ needed in this proof is its precongruence, which is
indeed true for all weakly guarded contexts: % removed "direct sums"
\begin{lemma}
\label{l:ruptocon}
Let $\til P$ and $\til Q$ be solutions (for $\rcontr$)
 of a system of weakly guarded contractions.
For any context $\qct$, 
if  $\ct{\til P}\Arr{\mu}  R$,
 then there is a context $\qctp$
such that $R \mcontrBIS \ctp{\til P}$ and  $\ct{\til Q} \Arr{\mu} R'
 \wb \ctp{\til Q}$ for some $R'$.
\end{lemma}

>>>>

For this new relation, the analogous of Lemma~\ref{l:uptocon} and of
Theorem~\ref{t:contraBisimulationU} can now be stated without
constraints on summation.
The schema of the proofs is almost identical, because
the only property % I&C editor issue 5
of $\rcontr$ needed in this proof is its precongruence, which is
indeed true for all weakly guarded contexts: % removed "direct sums"
\begin{lemma}
\label{l:ruptocon}
Let $\til P$ and $\til Q$ be solutions (for $\rcontr$)
 of a system of weakly guarded contractions.
For any context $\qct$, 
if  $\ct{\til P}\Arr{\mu}  R$,
 then there is a context $\qctp$
such that $R \mcontrBIS \ctp{\til P}$ and  $\ct{\til Q} \Arr{\mu} R'
 \wb \ctp{\til Q}$ for some $R'$.
\end{lemma}
